---
title: "Round-trip simulation example with bnlearn"
author: "Bob Horton"
date: "3/4/2021"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```

Note: This document makes use of packages that must be installed from BioConductor, including some dependencies that must be compiled. Assuming you already have `RTools`, you can install RGraphviz and dependencies as follows:
```
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install()
BiocManager::install(c("graph", "Rgraphviz"))

```
```{r libraries, warning=FALSE}
library(dplyr)
library(tidyr)
library(bnlearn)
```

This example follows Figure 2.1 in  [_Bayesian Artificial Intelligence_](https://bayesian-intelligence.com/publications/bai/), [chapter 2[(https://bayesian-intelligence.com/publications/bai/book/BAI_Chapter2.pdf).


We'll use the probabilities given in the figure to generate some simulated data, then see if `bnlearn` can determine the structure of the Bayes Net from the simulated data.

```{r cancer_simdata}
set.seed(42)

sim_cancer <- function(N){
  
  cancer_prob <- function(P, S){
    case_when(
      P=='H' & S=='T'  ~ 0.05,
      P=='H' & S=='F'  ~ 0.02,
      P=='L' & S=='T'  ~ 0.03,
      P=='L' & S=='F'  ~ 0.001
    )
  }
  
  xray_prob <- function(C) ifelse(C=='T', 0.9, 0.2)
  
  dyspnoea_prob = function(C) ifelse(C=='T', 0.65, 0.30)
  
  data.frame(
    Pollution = ifelse(runif(N) < 0.9, 'L', 'H'),
    Smoking = ifelse(runif(N) < 0.3, 'T', 'F'),
    LeftHanded = ifelse(runif(N) < 0.1, 'T', 'F'),
    RedHead = ifelse(runif(N) < 0.02, 'T', 'F')
  ) %>% mutate(
    Cancer = ifelse(cancer_prob(Pollution, Smoking) > runif(N), 'T', 'F'),
    Xray = ifelse(xray_prob(Cancer) > runif(N), 'T', 'F'),
    Dyspnoea = ifelse(dyspnoea_prob(Cancer) > runif(N), 'T', 'F')
  ) %>% lapply(factor) %>% as.data.frame
}

simdata <- sim_cancer(20000)

head(simdata)

```

Learn the structure:

```{r plot_gs}

cancer_structure <- simdata %>% fast.iamb # this is one of several structure-learning algorithms

modelstring(cancer_structure)

cancer_structure %>% graphviz.plot

# Try different numbers of training examples; more data usually works better.
# sim_cancer(10000) %>% fast.iamb %>% plot
```


## Fitting the network parameters

```{r fitting_parameters}

cancer_fit <- bn.fit(cancer_structure, data=simdata)

cancer_fit

```

## Making predictions

```{r making_predictions}

cpquery(cancer_fit, (Cancer=='T'), (Dyspnoea=='T' & Xray=='T' & Pollution=='H' & Smoking=='T'))

cpquery(cancer_fit, (Cancer=='T'), (Dyspnoea=='F' & Xray=='F' & Pollution=='H' & Smoking=='T'))

cpquery(cancer_fit, (Cancer=='T'), (Dyspnoea=='T' & Xray=='T'))

cpquery(cancer_fit, (Cancer=='T'), (Dyspnoea=='F' & Xray=='F'))
        
cpquery(cancer_fit, (Smoking=='T'), (Dyspnoea=='F' & Xray=='F'))
        
```
# Extracting Conditional Probability Tables

```{r extracting_CPTs}

reformat_CPT <- function(bnet, node_name){
  p_table <- bnet[[node_name]]$prob
  if ( (p_table %>% dim %>% length) == 1){
    prob_df <- p_table %>% as.matrix %>% t %>% as.data.frame
    names(prob_df) <- paste(node_name, names(prob_df), sep='_')
  } else {
    prob_df <- p_table %>% 
      ftable %>% 
      as.data.frame %>% 
      pivot_wider(names_from=node_name, values_from=Freq)
    output_cols <- c(ncol(prob_df) - 1, ncol(prob_df))
    names(prob_df)[output_cols] <- paste(node_name, names(prob_df)[output_cols], sep='_')
  }
  return(prob_df)
}


cancer_fit %>% reformat_CPT('Pollution')

cancer_fit %>% reformat_CPT('Smoking')

cancer_fit %>% reformat_CPT('Cancer')

cancer_fit %>% reformat_CPT('Xray')

cancer_fit %>% reformat_CPT('Dyspnoea')

```
